Rice yield improvements through plant breeding are offset by inherent yield declines over time

### 2017-10-29 output/breed_paper.Rnw
Updated and revised for grammar, etc.

Meeting the challenge of feeding a growing population with
increasingly limited resources will require increasing the yield
potential of staple crops, such as rice. Yet many high-yielding,
intensive production systems have experienced slow rates of yield
improvement in recent years despite a demonstrated increase in the
yield potential of new crop cultivars. We analyzed experimental data
from one such cropping system, i.e., California (CA) rice, in order to
quantify improvements made in the genetic yield potential obtained
through plant breeding. California rice systems are among the highest
in the world and close to maximum yield potential. Specifically, the
hypothesis was tested that if rice cultivar yields decline over time
(termed yield erosion) then apparent yield increases in side-by-side
yield comparison tests will not reflect increases in yield
potential. This hypothesis was tested using experimental yield data
over 30+ years at the California Cooperative Rice Research Foundation
Rice Experiment Station. Results show an apparent increase in yield in
side-by-side comparisons between old and new rice varieties was
observed when the yield erosion was not considered. However, the
yields of older varieties eroded at an estimated rate of 29.3 kg
ha\tsups{-1} year\tsups{-1} (90\% credible interval -4.4 to -53.3)
after initial selection. Once this effect was considered, the yield
advantage of newer cultivars over old was uncertain (-3.3 kg
ha\tsups{-1} year\tsups{-1}, 90\% credible interval -36.1 to
31.5). These results highlight the importance of continuous crop
improvement and deployment of new cultivars as yield will decline over
time without it. To increase the genetic yield potential of rice in
the face of this phenomenon, higher yield targets are needed to
counter-act the decrease over time in the current cultivars.

@@

Constraints on arable land are increasing simultaneous with the need
to increase total food production to meet a growing demand
\citep{Foley2011a, Godfray2010, Mueller2012}, which has necessitated
harvesting more grain per unit land area \citep{Lobell2009,
Tittonell2014}. Historically, agricultural research has been
successful in staving off the ``Malthusian catastrophe'' of demand
surpassing supply via continued yield improvement of staple
crops. However, many production systems are experiencing plateaus in
grain yield \citep{Grassini2013, Peng2010}. If this trend continues,
improvements per unit land area are no longer possible, and an
increase in the area under cultivation will be needed to meet food
demand, which carries with it undesirable ecological
implications~\citep{Foley2011a, tilman1999}. Therefore, it is of
critical importance to better understand why the rate of increase in
grain yields has declined or leveled off in intensified production
systems.

California (CA) rice represents one such production system. Rice is
grown primarily in the Sacramento Valley which is characterized by
having a Mediterranean climate, with long days, dry growing season,
and relatively free of pests and diseases, which lead to some of the
highest yields in the world~\citep{FAO2016}. Production in CA is
predominately focused on premium quality medium grain
\textit{japonica} cultivars (e.g., CalRose rice), and rice from CA is
recognized globally for its quality
(\url{http://agfax.com/2015/11/03/rice-calrose-wins-best-rice-world-competition/}). Most
cultivars in use in CA are developed by the California Cooperative
Rice Research Foundation (CCRRF; a collaboration between the
University of California, the USDA-Agricultural Research Service, and
farmer-funded research). In part due to improvements in rice genetics,
rice yields in CA increased rapidly during the period from 1920 --
1990; however, since the 1990s, the rate of yield increase has slowed
(Fig. \ref{fig:state_yield}) despite continuous crop
improvement. Based on a yield-gap analysis, on-farm yields in the
major rice growing region of CA is 73 to 76\% of maximum yield
potential~\citep{espe16a}. In highly intensive systems such as this,
\citet{grassini2011} have shown that farmers are capable of attaining
85\% of the maximum yield potential. Thus, given that farmers are near
the attainable yield potential, increases in yield are likely to be
slow and \citet{espe16a} reported that indeed this was the case
with yields on average increasing by about 50 to 62 kg ha\tsups{-1} between
1999 and 2014.

Certainly one question in looking at this situation is, ``do new
varieties have increased yield potential''?  Like many breeding
programs, one of the goals of the CCRRF is to increase yields through
breeding varieties with increased yield potential. However, when the
popular cultivar M-202 was released in 1985, it was reported to yield
an average of 11 Mg ha\tsups{-1} \citep[14\% moisture content
(MC);][]{m202}. In 2015, the reported average yield of the newest
released cultivar M-209 was 10.8 Mg ha\tsups{-1} \citep[14\%
MC;][]{m209}. However, in annual breeding reports and varietal release
announcements, based on yield performance in side-by-side variety
trails, data show that new varieties usually out yield the older
varieties. In other rice systems, a decrease in a cultivar's yield
performance over time has been observed \citep{Peng1998, Peng2010,
Peng2000, Datta1995}, often termed yield erosion \citep{Peng1998}. It
has been speculated that this yield erosion is due to the inability of
a cultivar to adapt to changing biotic and abiotic
conditions~\citep{Peng1998, Peng2010, Peng2000, Datta1995}. Yield
erosion may explain the apparent contradiction above, in which
side-by-side yield comparisons show yield improvements in varieties
while overall there may be little change in yield potential. To
address this issue we tested the hypothesis that if rice cultivar
yields erode over time then apparent yield increases in side-by-side
yield comparison tests will not reflect true yield increases.o

@@

The CCRRF Rice Experiment Station (RES) is located near Biggs, CA
(39.4648, -121.7342), and has been the central location for CCRRF's
efforts to develop improved rice cultivars adapted to CA since
1969. The California Statewide Variety Trails evaluates current and
promising varieties at the RES and six to eight on-farm trials around
the state each year. To avoid potential complications due to site,
climate, and management differences at the farmer-managed trials per
year, we focused our investigation on the RES. The RES is
researcher-managed and plants the majority of the experimental plots
for the statewide program (approximately 3 to 4 times the number of
experimental plots compared to other on-farm locations). The climate
at the RES is Mediterranean, characterized by mild winters during
which most of the annual precipitation occurs (\Sexpr{18*24.5} to
\Sexpr{25*24.5} mm annually) and warm summers largely free of
precipitation events. Soils at the RES are classified as
Esquon-Neerdobe clays with roughly 2\% soil organic matter
\citep{ssurgo} in the top 15cm.

As part of the annual Statewide Variety Trials, newly developed
entries are tested against officially released cultivars (checks) at
the RES in several trials spanning multiple planting dates per
year. For the purposes of this study, we concentrated on released
medium grain cultivars, as these cultivars constitute approximately
90\% of the planted rice area in CA. Newly developed entries are typically
evaluated in these trials for 3 to 5 years prior to
their official release. For these cultivars planted in breeding trials
at the RES from \Sexpr{min(mod_data[,'year'])} --
\Sexpr{max(mod_data[,'year'])} plot level observations (replicated
3--4 times in a CRB design) were collected leading to a final data set
of \Sexpr{format(nrow(mod_data), big.mark = ',')} observations
representing \Sexpr{length(unique(mod_data[,'id']))} cultivars and
included every publicly released medium grain cultivar developed by
the CCRRF from \Sexpr{min(mod_data[,'release_yr'])} --
\Sexpr{max(mod_data[,'release_yr'])}. For each cultivar, the official
year of public release and the time elapsed (in years) since release was
determined.

Experimental plots at the RES are managed similar to University of California
Cooperative Extension following prescribed best management
practice for land preparation, fertility rates, and pest
and disease control \citep{ucce}. As is typical for water-seeded
systems, for planting, rice seed is pre-germinated prior
to direct-seeding into pre-flooded fields. Plots were either 14.0 or
18.5 m\tsups{2}, and were harvested after physiological
maturity using a small-plot combine. All yields were converted to 14\%
moisture content prior to reporting.

@@

Data were processed and models were fit in R, an environment for
statistical computing \citep{R}. Models were fit using the `rstanarm'
package \citep{rstanarm}, an interface to Stan, a language for
probabilistic programming \citep{stan}. All model diagnostics,
including $\hat{R}$, effective size, and posterior predictive checks
were examined before reporting results. To test goodness of fit, the
full model was compared to models containing only the years since
release or the year of release using the `loo' package \citep{loo}, an
efficient means of conducting leave-one-out cross validation
\citep{loo_paper}. The 90\% credible interval (90\% CI), defined as
the interval containing 90\% of the distribution of estimated credible
parameter values given the data, was calculated as a measure of
uncertainty in the parameter estimates using the quantile method. The
90\% CI is more stable to perpetuation's and hence is preferable to the 95\%
interval \citep{rstanarm}. To verify the estimates, the analysis was
also conducted using classical methods (frequentist mixed-effects
model fit by maximum likelihood; See supplemental material for
details). Complete data and code used for this
analysis is available at \url{github.com/mespe/rice_advances.git}.

\section{Results}\label{results}
% Invert this section, then talk about yield improvement accounting
% for yield erosion.

@@

There was agreement between the estimates from the Bayesian and
classical approaches.  The Bayesian estimates are presented here,
though the classical estimates can be found in the supplemental
material.  Comparing the cultivars' yield during each's initial time
(average 3 years) in the experimental trials just before being
released showed that across cultivars, yields ranged from
\Sexpr{round(plot_data["M207", "mean"], 2)} (M-207) to
\Sexpr{round(plot_data["M205", "mean"], 2)} Mg ha\tsups{-1}
(M-205). Evaluating these initial yields over time indicates there was
no evidence of genetic yield improvements during the 30-year study
period (Fig. 2). Individual exceptions to this were two varieties
(M-204 and M-205) that had higher yields than the others.

@@

The yields of each cultivar were evaluated over time relative to the
time when the cultivar was released using the full model which
accounts for both yield erosion and yield improvements over time. The
average decrease in yield over time (i.e., yield erosion) once a
cultivar was selected was estimated \Sexpr{yrs_coef} kg ha\tsups{-1}
year\tsups{-1} (90\% CI: \Sexpr{RES_CI[2]} to \Sexpr{RES_CI[1]}) (Fig
\ref{fig:RES_year}; Table \ref{tbl:coef}). The oldest cultivar in the
trials (\Sexpr{unique(mod_data[which.min(mod_data[,"year"]),"id"])}
had lost an average of \Sexpr{round(yrs_coef*-36/1000,1)} Mg
ha\tsups{-1} since its release. The most dominate medium grain rice
cultivar, M-202 (widely planted from the late 1980s until the early
2000s) experienced similar trends in yield decline
(Fig. \ref{fig:RES_year}; Table \ref{tbl:coef}). M-206 replaced M-202
as the most widely grown cultivar in CA and the current estimated 560
kg ha-1 yield advantage of M-206 over M-202 can be attributed solely
to the erosion of M-202's yield over time (Figs \ref{fig:first_yrs} \&
\ref{fig:compare_yrs}). Further supporting this estimate of yield
erosion, both cultivars yielded similarly at their respective year of
release (Fig. \ref{fig:compare_yrs}), despite the fact that M-202 was
released 19 years prior to M-206 (Fig. \ref{fig:compare_yrs}, Table
\ref{tbl:ids}).

When yield erosion was not corrected for in the analysis (i.e.,
omitted from the model), there is an apparent small increase per year
(approximately \Sexpr{only_release_coef} kg ha\tsups{-1}
year\tsups{-1}; 90\% CI \Sexpr{only_release_CI[1]} to
\Sexpr{only_release_CI[2]}) (Fig. \ref{fig:wo_erosion}).  However,
using the full model which corrected for both the year of release and
yield erosion, indicates that the estimated change in grain yield per
year in the CCRRF's breeding trials at the RES was
\Sexpr{release_yr_coef} kg ha\tsups{-1} year\tsups{-1}, with both
small positive and negative trends within the credible interval (90\%
CI: \Sexpr{rel_CI[1]} to \Sexpr{rel_CI[2]})
(Fig. \ref{fig:first_yrs}).

@@

In California, the cultivars with broad adaptation and adoption over
the 30+ year period of this study are M-202, M-206, and M-105. Cold
tolerance is a major problem for much of the California rice producing
region ~\citep{espe2017}, although it is a bigger problem in the
southern portion of the Sacramento Valley due to the influence of the
ocean winds through the Delta ~\citep{espe16a}. These three
cultivars have a higher degree of cold tolerance, high yield
potential, and relatively short duration and are thus broadly grown
with over 70\% of the area in one of these cultivars during the study
period ~citep{ucce}. Importantly, not all medium grain cultivars that were
released were purported to have higher yields than previous
cultivars. Some cultivars had special traits, but upon release had
similar yields to more broadly adapted cultivars. Examples of this
include M-208, which had resistance to blast disease (scientific name)
and M-402 which is a premium quality variety. Other varieties such as
M-205 and M-209, while high yielding, do not have the cold tolerance
of the other cultivars and are thus only recommended for the northern
portion of the Sacramento Valley.

Looking at all the cultivars that have been released, there is no
conclusive evidence of increased genetic yield potential in medium
grain rice cultivars released by CCRRF's breeding improvement program
over the period \Sexpr{min(mod_data[,'year'])} to
\Sexpr{max(mod_data[,'year'])}, which has stayed relatively constant
at \Sexpr{round(mean(mod_data[,"yield_kg"]), 2)} Mg
ha\tsups{-1}. Since the estimated yield improvement spans from
\Sexpr{rel_CI[1]} to \Sexpr{rel_CI[2]} kg ha\tsups{-1} year\tsups{-1}
(Fig. \ref{fig:RES_year}; Table \ref{tbl:coef}),
both increases and decreases in the genetic yield potential over the
study period are plausible. The estimated yield
improvement of \Sexpr{only_release_coef} kg ha\tsups{-1}
year\tsups{-1} (Fig. \ref{fig:wo_erosion}) that does not account for
yield erosion is lower than the 37 to 60 kg ha\tsups{-1}
year\tsups{-1} estimated by~\citet{Omar2014} in side-by-side
comparisons. Rather, this analysis suggests that the CCRRF breeding
program has maintained but not substantially improved the yield
potential in the face of changing biotic and abiotic pressures. This
conclusion holds true even when evaluating the cultivars which have
been breed for broad adaption and high yield potential (M-202, M-206
and M-105). Yields of these varieties at time of release have all
yielded between
\Sexpr{tmp=mod_data[mod_data[,"yrs_in_trial"]<=0&mod_data[,"id"]%in%c("M202","M206","M105"),];round(min(by(tmp,tmp[,"id"],function(x)mean(x[,"yield_kg"]))/1000),2)}
and
\Sexpr{round(min(by(tmp,tmp[,"id"],function(x)mean(x[,"yield_kg"]))/1000),2)}
Mg ha\tsups{-1} with no indication of yields increasing
(Fig. \ref{fig:first_yrs}). While these yields have not increased,
M-205 did have higher yield potential; however, this variety is not
broadly adapted due to poor cold tolerance.

Yields of a particular cultivar may decline overtime for a number of
reasons. First, ~\citet{Nie2009} reported yield declines in rice due
to continuously flooded conditions over a long time period which could
be overcome with improved N management. This is likely not the case
here as the fields used in this study grow rice only once season a
year and tend to be non-flooded for the remainder of the
year. Furthermore, at this location and unlike most CA rice fields,
the fields are left fallow every other year. Second, changing climate
may influence yields with either warmer~\citep{van2013} or
cooler (Espe et al., 2017) temperatures reducing yields. Again, this
is unlikely to be the case in this situation as Espe et al. (2017)
found no significant changes in growing season temperatures (either
high, low or average temps) the Sacramento Valley during the study
period.  Thirdly, yield erosion of a cultivar as has been reported for
the cultivar IR8 at IRRI~\citep{Peng1998, Peng2000, Peng2010}. Biotic
and abiotic forces have previously been speculated to be responsible
for yield erosion, since the cultivar's genetics remain stable from
the time of selection but the conditions the cultivar may experience
over time are continuously changing~\citep{Peng1998, Peng2000,
Peng2010}. We estimate that yield erosion has decreased the yield of
medium grain rice cultivars released by the CCRRF over the last 30+
years by an average of \Sexpr{yrs_coef} Mg ha\tsups{-1}, potentially
offsetting any gains made by the crop improvement program. However, it
is not clear what is responsible for this erosion. It is unlikely
biotic factors as pest and disease pressure are low in this
Mediterranean climate. Furthermore, these fields are fallowed every
other year and the residues are burned, further lowering the potential
for pest and disease pressure. While efforts should be made to
identify the cause of erosion in this environment, this observation
highlights both the need for maintenance of plant breeding and the
difficulty of increasing genetic yield potential, especially in
systems such as CA where yield is approaching the physiological limit
\citep{espe16a}. Without maintenance breeding (i.e., continuously
selecting cultivars well adapted to current conditions), CA rice
production systems are predicted to have lost an estimated
\Sexpr{round(yrs_coef*(-36/1000),1)} Mg ha\tsups{-1} grain yield over
the study period.

A goal of many breeding programs is to not only maintain the benchmark
yield of the previous ``standard'' but to surpass it, especially if we
are to meet the target of 50\% increased food production by 2050
\citep{Foley2011a}. Given the estimated yield erosion, to hit an
ambitious target of 0.5 Mg ha\tsups{-1} (approximately 5\%) increase
in yield per release (with roughly 4 years between releases), a new
cultivar needs to yield an additional \Sexpr{round(yrs_coef*-4,-1)} kg
ha\tsups{-1} in side-by-side comparison in addition to the 0.5 Mg
ha\tsups{-1} target improvement. Often side-by-side yield improvements
show an improvement in grain yield of 2\% per release, in which case
no absolute improvements in yield have been made beyond maintaining
yield. The evidence here is consistent with the observation that the
genetic yield potential has not increased substantially in the last
30+ years (Fig. \ref{fig:first_yrs}).

>>>>

Meeting the challenge of feeding a growing population with
increasingly limited resources will require increasing the yield
potential of staple crops, such as rice. Yet many high-yielding,
intensive production systems have experienced slow rates of yield
improvement in recent years despite a demonstrated increase in the
yield potential of new crop cultivars. We analyzed experimental data
from one such cropping system, i.e., California (CA) rice, in order to
quantify improvements made in the genetic yield potential obtained
through plant breeding. California rice systems are among the highest
in the world and close to maximum yield potential. Specifically, the
hypothesis was tested that if rice cultivar yields decline over time
(termed yield erosion) then apparent yield increases in side-by-side
yield comparison tests will not reflect increases in yield
potential. This hypothesis was tested using experimental yield data
over 30+ years at the California Cooperative Rice Research Foundation
Rice Experiment Station. Results show an apparent increase in yield in
side-by-side comparisons between old and new rice varieties when the
yield erosion was not considered. However, the yields of older
varieties eroded at an estimated rate of 29.3 kg ha\tsups{-1}
year\tsups{-1} (90\% credible interval -4.4 to -53.3) after initial
selection. Once this effect was considered, the yield advantage of
newer cultivars over old was uncertain (-3.3 kg ha\tsups{-1}
year\tsups{-1}, 90\% credible interval -36.1 to 31.5). This suggests
that side-by-side comparisons reflected the decreasing yields of older
cultivars more than the increasing yield of newer ones. These results
highlight the importance of continuous crop improvement and deployment
of new cultivars as yield will decline over time without it. To
increase the genetic yield potential of rice in the face of this
phenomenon, higher yield targets are needed to counter-act the
decrease over time in the current cultivars.

@@

Constraints on arable land are increasing simultaneous with the need
to increase total food production to meet a growing demand
\citep{Foley2011a, Godfray2010, Mueller2012}, which has necessitated
harvesting more grain per unit land area~\citep{Lobell2009,
Tittonell2014}. Historically, agricultural research has been
successful in staving off the ``Malthusian catastrophe'' of demand
surpassing supply via continued yield improvement of staple
crops. However, many production systems are experiencing plateaus in
grain yield~\citep{Grassini2013, Peng2010}. If this trend continues,
improvements per unit land area are no longer possible, and an
increase in the area under cultivation will be needed to meet food
demand, which carries with it undesirable ecological
implications~\citep{Foley2011a, tilman1999}. Therefore, it is of
critical importance to better understand why the rate of increase in
grain yields has declined or leveled off in intensified production
systems.

California (CA) rice represents one such production system. Rice is
grown primarily in the Sacramento Valley, which is characterized by
having a Mediterranean climate with long days, a dry growing season,
and relatively free of pests and diseases. These conditions lead to
some of the highest yields in the world~\citep{FAO2016}. Production in
CA is predominately focused on premium quality medium grain
\textit{japonica} cultivars (e.g., CalRose rice), and rice from CA is
recognized globally for its quality
(\url{http://agfax.com/2015/11/03/rice-calrose-wins-best-rice-world-competition/}). Most
cultivars in use in CA are developed by the California Cooperative
Rice Research Foundation (CCRRF; a collaboration between the
University of California, the USDA-Agricultural Research Service, and
farmer-funded research). In part due to improvements in rice genetics,
rice yields in CA increased rapidly during the period from 1920 --
1990; however, since the 1990s, the rate of yield increase has slowed
(Fig. \ref{fig:state_yield}) despite continuous crop
improvement. Based on a yield-gap analysis, on-farm yields in the
major rice growing region of CA is 73 to 76\% of maximum yield
potential~\citep{espe16a}. In highly intensive systems such as this,
\citet{grassini2011} have shown that farmers are capable of attaining
85\% of the maximum yield potential. Thus, given that farmers are near
the attainable yield potential, increases in yield are likely to be
slow and \citet{espe16a} reported that indeed this was the case
with yields on average increasing by about 50 to 62 kg ha\tsups{-1}
year\tsups{-1} between 1999 and 2014.

Certainly one question in looking at this situation is, ``do new
varieties have increased yield potential''?  Like many breeding
programs, one of the goals of the CCRRF is to increase yields through
breeding varieties with increased yield potential. However, when the
popular cultivar M-202 was released in 1985, it was reported to yield
an average of 11 Mg ha\tsups{-1}~\citep[14\% moisture content
(MC);][]{m202}. In 2015, the reported average yield of the newest
released cultivar M-209 was 10.8 Mg ha\tsups{-1}~\citep[14\%
MC;][]{m209}. However, in annual breeding reports and varietal release
announcements, data based on yield performance in side-by-side variety
trails show that new varieties usually out yield the older
varieties. In other rice systems, a decrease in a cultivar's yield
performance over time has been observed~\citep{Peng1998, Peng2010,
Peng2000, Datta1995}, often termed yield erosion~\citep{Peng1998}. It
has been speculated that this yield erosion is due to the inability of
a cultivar to adapt to changing biotic and abiotic
conditions~\citep{Peng1998, Peng2010, Peng2000, Datta1995}. Yield
erosion may explain the apparent contradiction above, in which
side-by-side yield comparisons show yield improvements in varieties
while overall there may be little change in yield potential. To
address this issue we tested the hypothesis that if rice cultivar
yields erode over time then apparent yield increases in side-by-side
yield comparison tests will not reflect true or absolute yield
increases over time.

@@

The CCRRF Rice Experiment Station (RES) is located near Biggs, CA
(39.4648, -121.7342), and has been the central location for CCRRF's
efforts to develop improved rice cultivars adapted to CA since
1969. The California Statewide Variety Trails evaluate current and
promising varieties at the RES and six to eight on-farm trials around
the state each year. To avoid potential complications due to site,
climate, and management differences at the farmer-managed trials per
year, we focused our investigation on the RES. The RES is
researcher-managed and plants the majority of the experimental plots
for the statewide program (approximately 3 to 4 times the number of
experimental plots compared to other on-farm locations). The climate
at the RES is Mediterranean, characterized by mild winters during
which most of the annual precipitation occurs (\Sexpr{18*24.5} to
\Sexpr{25*24.5} mm annually) and warm summers largely free of
precipitation events. Soils at the RES are classified as
Esquon-Neerdobe clays with roughly 2\% soil organic matter in the top 15cm~\citep{ssurgo}.

As part of the annual Statewide Variety Trials, newly developed
entries are tested against officially released cultivars (checks) at
the RES in several trials spanning multiple planting dates per
year. For the purposes of this study, we concentrated on released
medium grain cultivars, as these cultivars constitute approximately
90\% of the planted rice area in CA. Newly developed entries are typically
evaluated in these trials for 3 to 5 years prior to
their official release. For these cultivars planted in breeding trials
at the RES from \Sexpr{min(mod_data[,'year'])} --
\Sexpr{max(mod_data[,'year'])} plot level observations (replicated
3 to 4 times in a CRB design) were collected leading to a final data set
of \Sexpr{format(nrow(mod_data), big.mark = ',')} observations
representing \Sexpr{length(unique(mod_data[,'id']))} cultivars and
included every publicly released medium grain cultivar developed by
the CCRRF from \Sexpr{min(mod_data[,'release_yr'])} --
\Sexpr{max(mod_data[,'release_yr'])}. For each cultivar, the official
year of public release and the time elapsed (in years) since release was
determined.

Experimental plots at the RES are managed similar to University of California
Cooperative Extension prescribed best management
practice for land preparation, fertility rates, and pest
and disease control~\citep{ucce}. As is typical for water-seeded
systems, the plots were planted by pre-germinating rice seed prior
to direct-seeding into pre-flooded fields. Plots were either 14.0 or
18.5 m\tsups{2}, and were harvested after physiological
maturity using a small-plot combine. All yields were converted to 14\%
MC prior to reporting.

@@

Data were processed and models were fit in R, an environment for
statistical computing~\citep{R}. Models were fit using the `rstanarm'
package~\citep{rstanarm}, an interface to Stan, a language for
probabilistic programming~\citep{stan}. All model diagnostics,
including $\hat{R}$, effective size, and posterior predictive checks
were examined before reporting results. To test goodness of fit, the
full model was compared to models containing only the years since
release or the year of release using the `loo' package~\citep{loo}, an
efficient means of conducting leave-one-out cross validation
\citep{loo_paper}. The 90\% credible interval (90\% CI), defined as
the interval containing 90\% of the distribution of estimated credible
parameter values given the data, was calculated as a measure of
uncertainty in the parameter estimates using the quantile method. The
90\% CI is more stable to perpetuation's and hence is preferable to the 95\%
interval~\citep{rstanarm}. To verify the estimates, the analysis was
also conducted using classical methods (frequentist mixed-effects
model fit by maximum likelihood; See supplemental material for
details). Complete data and code used for this
analysis is available at \url{github.com/mespe/rice_advances.git}.

\section{Results}\label{results}

@@

There was general agreement between the estimates from the Bayesian and
classical approaches.  The Bayesian estimates are presented here due
to the inherent flexibility of Bayesian inference, 
though the classical estimates can be found in the supplemental
material. Comparing the cultivars' yield during each's initial time
(average 3 years) in the experimental trials just before being
released showed that across cultivars, yields ranged from
\Sexpr{round(plot_data["M207", "mean"], 2)} (M-207) to
\Sexpr{round(plot_data["M205", "mean"], 2)} Mg ha\tsups{-1}
(M-205). Evaluating these initial yields over time indicates there was
no evidence of genetic yield improvements during the 30-year study
period (Fig. 2). Individual exceptions to this were two varieties
(M-204 and M-205) that had higher yields than the others.

@@

The yields of each cultivar were evaluated over time relative to the
time when the cultivar was released using the full model which
accounts for both yield erosion and yield improvements over time. The
average decrease in yield over time (i.e., yield erosion) once a
cultivar was selected was estimated \Sexpr{yrs_coef} kg ha\tsups{-1}
year\tsups{-1} (90\% CI: \Sexpr{RES_CI[2]} to \Sexpr{RES_CI[1]}) (Fig
\ref{fig:RES_year}; Table \ref{tbl:coef}). The oldest cultivar in the
trials (\Sexpr{gsub("M","M-",unique(mod_data[which.min(mod_data[,"release_yr"]),"id"]))})
had lost an average of \Sexpr{round(yrs_coef*-36/1000,1)} Mg
ha\tsups{-1} since its release. The most dominate medium grain rice
cultivar, M-202 (widely planted from the late 1980s until the early
2000s) experienced similar trends in yield decline
(Fig. \ref{fig:RES_year}; Table \ref{tbl:coef}). M-206 replaced M-202
as the most widely grown cultivar in CA and the current estimated 560
kg ha\tsups{-1} yield advantage of M-206 over M-202 can be attributed solely
to the erosion of M-202's yield over time (Figs \ref{fig:first_yrs} \&
\ref{fig:compare_yrs}). Further supporting this estimate of yield
erosion, both cultivars yielded similarly at their respective year of
release (Fig. \ref{fig:compare_yrs}), despite the fact that M-202 was
released 19 years prior to M-206 (Fig. \ref{fig:compare_yrs}; Table
\ref{tbl:ids}).

When yield erosion is not corrected for in the analysis (i.e.,
omitted from the model), there is an apparent small increase per year
(approximately \Sexpr{only_release_coef} kg ha\tsups{-1}
year\tsups{-1}; 90\% CI \Sexpr{only_release_CI[1]} to
\Sexpr{only_release_CI[2]}) (Fig. \ref{fig:wo_erosion}).  However,
using the full model which corrected for both the year of release and
yield erosion, indicates that the estimated change in grain yield per
year in the CCRRF's breeding trials at the RES was
\Sexpr{release_yr_coef} kg ha\tsups{-1} year\tsups{-1}, with both
small positive and negative trends within the credible interval (90\%
CI: \Sexpr{rel_CI[1]} to \Sexpr{rel_CI[2]})
(Fig. \ref{fig:first_yrs}).

@@

In California, the cultivars with broad adaptation and adoption over
the 30+ year period of this study are M-202, M-206, and M-105. Cold
tolerance is a major problem for much of the California rice producing
region ~\citep{espe2017}, although it is a bigger problem in the
southern portion of the Sacramento Valley due to the influence of the
ocean winds through the Delta ~\citep{espe16a}. These three
cultivars have a higher degree of cold tolerance, high yield
potential, and relatively short duration and are thus broadly grown
with over 70\% of the area in one of these cultivars during the study
period~\citep{ucce}. Importantly, not all medium grain cultivars that were
released were purported to have higher yields than previous
cultivars. Some cultivars had special traits, but upon release had
similar yields to more broadly adapted cultivars. Examples of this
include M-208, which had resistance to blast disease (\textit{Magnaporthe grisea})
and M-402 which is a premium quality variety. Other varieties such as
M-205 and M-209, while high yielding, do not have the cold tolerance
of the other cultivars and are thus only recommended for the northern
portion of the Sacramento Valley.

Looking at all the cultivars that have been released, there is no
conclusive evidence of increased genetic yield potential in medium
grain rice cultivars released by CCRRF's breeding improvement program
over the period \Sexpr{min(mod_data[,'year'])} to
\Sexpr{max(mod_data[,'year'])}, which has stayed relatively constant
at \Sexpr{round(mean(mod_data[,"yield_kg"])/1000, 1)} Mg
ha\tsups{-1}. Since the estimated yield improvement spans from
\Sexpr{rel_CI[1]} to \Sexpr{rel_CI[2]} kg ha\tsups{-1} year\tsups{-1}
(Fig. \ref{fig:RES_year}; Table \ref{tbl:coef}),
both increases and decreases in the genetic yield potential over the
study period are plausible. The estimated yield
improvement of \Sexpr{only_release_coef} kg ha\tsups{-1}
year\tsups{-1} (Fig. \ref{fig:wo_erosion}) that does not account for
yield erosion is lower than the 37 to 60 kg ha\tsups{-1}
year\tsups{-1} estimated by~\citet{Omar2014} in side-by-side
comparisons. Rather, this analysis suggests that the CCRRF breeding
program has maintained but not substantially improved the yield
potential in the face of changing biotic and abiotic pressures. This
conclusion holds true even when evaluating the cultivars which have
been breed for broad adaption and high yield potential (M-202, M-206
and M-105). Yields of these varieties at time of release have all
yielded between
\Sexpr{tmp=mod_data[mod_data[,"yrs_in_trial"]<=0&mod_data[,"id"]%in%c("M202","M206","M105"),];round(min(by(tmp,tmp[,"id"],function(x)mean(x[,"yield_kg"]))/1000),2)}
and
\Sexpr{round(min(by(tmp,tmp[,"id"],function(x)mean(x[,"yield_kg"]))/1000),2)}
Mg ha\tsups{-1} with no indication of yields increasing
(Fig. \ref{fig:first_yrs}). While these yields have not increased,
M-205 did have higher yield potential; however, this variety is not
broadly adapted due to poor cold tolerance.

Yields of a particular cultivar may decline overtime for a number of
reasons. First, ~\citet{Nie2009} reported yield declines in rice due
to continuously flooded conditions over a long time period which could
be overcome with improved N management. This is likely not the case
here as the fields used in this study grow rice only once season a
year and tend to be non-flooded for the remainder of the
year. Furthermore, at this location and unlike most CA rice fields,
the fields are left fallow every other year. Second, changing climate
may influence yields with either warmer~\citep{van2013} or
cooler (Espe et al., 2017) temperatures reducing yields. Again, this
is unlikely to be the case in this situation as Espe et al. (2017)
found no significant changes in growing season temperatures (either
high, low or average temps) the Sacramento Valley during the study
period.  Thirdly, yield erosion of a cultivar as has been reported for
the cultivar IR8 at IRRI~\citep{Peng1998, Peng2000, Peng2010}. Biotic
and abiotic forces have previously been speculated to be responsible
for yield erosion, since the cultivar's genetics remain stable from
the time of selection but the conditions the cultivar may experience
over time are continuously changing~\citep{Peng1998, Peng2000,
Peng2010}. We estimate that yield erosion has decreased the yield of
medium grain rice cultivars released by the CCRRF over the last 30+
years by an average of \Sexpr{yrs_coef} Mg ha\tsups{-1} year\tsups{-1}, potentially
offsetting any gains made by the crop improvement program. However, it
is not clear what is responsible for this erosion. It is unlikely
biotic factors as pest and disease pressure are low in this
Mediterranean climate. Furthermore, these fields are fallowed every
other year and the residues are burned, further lowering the potential
for pest and disease pressure. While efforts should be made to
identify the cause of erosion in this environment, this observation
highlights both the need for maintenance of plant breeding and the
difficulty of increasing genetic yield potential, especially in
systems such as CA where yield is approaching the physiological limit
\citep{espe16a}. Without maintenance breeding (i.e., continuously
selecting cultivars well adapted to current conditions), CA rice
production systems are predicted to have lost an estimated
\Sexpr{round(yrs_coef*(-36/1000),1)} Mg ha\tsups{-1} grain yield over
the study period.

There are several cultivars that were found to have better than
average yields, even after correcting for yield erosion. However,
these cultivars (M-201, M-204, M-205, and possibly M-209;
Fig. \ref{fig:ids_comp})) are not widely adopted. It is known that
cool temperatures during the booting stage can have a large impact on
grain yields~\citep{espe2017}, and the RES experiences warmer
temperatures than many other rice growing areas in CA
\citep{espe2017}. These cultivars may perform well at the RES but not
widely in CA. Therefore, it should be noted that while our
investigation has been concerned about increasing yield potential, the
estimates here might not capture yield stability. Yield stability is
notoriously difficult to quantify~\citep[e.g.,][]{lobell2011}, hence
further investigation is needed to explore how yield stability has
changed over the study period.

A goal of many breeding programs is to not only maintain the benchmark
yield of the previous ``standard'' but to surpass it, especially if we
are to meet the target of 50\% increased food production by 2050
\citep{Foley2011a}. Given the estimated yield erosion, to hit an
ambitious target of 500 kg ha\tsups{-1} (approximately 5\%) increase
in yield per release (with roughly 4 years between releases), a new
cultivar needs to yield an additional \Sexpr{round(yrs_coef*-4,-1)} kg
ha\tsups{-1} in side-by-side comparison in addition to the 500 kg
ha\tsups{-1} target improvement. Often side-by-side yield improvements
show an improvement in grain yield of 2 to 3\% per release, in which case
no absolute improvements in yield have been made beyond maintaining
yield. The evidence here is consistent with the observation that the
genetic yield potential has not increased substantially in the last
30+ years (Fig. \ref{fig:first_yrs}).
